---
title: "IPBES BM Fact Checker — NLI Scores: Basic Overview"
subtitle: "GA1 · config: deberta_zeroshot"
date: today
date-format: "D MMMM YYYY"
author: "IPBES Data Team"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: false
    embed-resources: true
    smooth-scroll: true
    theme: none
    css: |
      /* ── tokens ── */
      :root {
        --c-forest:   #1C3A2E;
        --c-moss:     #2E6B4F;
        --c-sienna:   #8B2E2E;
        --c-sand:     #A8896A;
        --c-parch:    #F5F3EE;
        --c-body:     #3A4A41;
        --c-muted:    #6B7B71;
        --c-rule:     #D4CFC6;
        --c-sup-bg:   #EBF3EE;
        --c-ref-bg:   #F5EBEB;
        --c-nei-bg:   #F5F1EB;
        --c-card:     #FFFFFF;
        --r-card:     6px;
        --font-serif: Georgia, "Times New Roman", serif;
        --font-ui:    system-ui, -apple-system, sans-serif;
        --font-mono:  "SF Mono", "Fira Code", monospace;
      }

      /* ── base ── */
      body {
        font-family:  var(--font-serif);
        font-size:    16px;
        line-height:  1.7;
        color:        var(--c-body);
        background:   var(--c-parch);
        margin: 0;
        padding: 0;
      }

      /* ── layout ── */
      #quarto-content, .page-layout-article {
        background: var(--c-parch);
      }
      .content-container, #quarto-margin-sidebar, main {
        max-width: 860px;
      }

      /* ── headings ── */
      h1, h2, h3, h4 {
        font-family: var(--font-ui);
        color:       var(--c-forest);
        font-weight: 700;
        text-wrap:   balance;
        line-height: 1.25;
      }
      h1 { font-size: 2rem;   border-bottom: 3px solid var(--c-moss);
           padding-bottom: .5rem; margin-top: 2rem; }
      h2 { font-size: 1.35rem; border-bottom: 1px solid var(--c-rule);
           padding-bottom: .3rem; margin-top: 2.5rem; }
      h3 { font-size: 1.1rem; color: var(--c-moss); margin-top: 2rem; }

      /* ── subtitle block ── */
      .quarto-title-meta { font-family: var(--font-ui); font-size: .85rem;
                            color: var(--c-muted); }
      .quarto-title-meta-heading { color: var(--c-forest); font-weight: 600; }

      /* ── tables ── */
      .table-wrap { overflow-x: auto; margin: 1.5rem 0; }
      table {
        font-family:  var(--font-ui);
        font-size:    .82rem;
        border-collapse: collapse;
        width: 100%;
        min-width: 480px;
        font-variant-numeric: tabular-nums;
      }
      thead tr { background: var(--c-forest); color: #fff; }
      thead th { padding: .5rem .75rem; text-align: left; font-weight: 600;
                  letter-spacing: .03em; }
      tbody tr:nth-child(odd)  { background: var(--c-card); }
      tbody tr:nth-child(even) { background: #EFEDE8; }
      tbody td { padding: .4rem .75rem; border-bottom: 1px solid var(--c-rule); }
      tbody tr:last-child td   { border-bottom: none; }

      /* ── callout boxes ── */
      .callout {
        border-left: 4px solid var(--c-moss);
        background:  var(--c-card);
        padding: 1rem 1.25rem;
        margin: 1.75rem 0;
        border-radius: 0 var(--r-card) var(--r-card) 0;
        box-shadow: 0 1px 4px rgba(0,0,0,.07);
      }
      .callout.warning { border-color: var(--c-sienna); background: var(--c-ref-bg); }
      .callout.neutral { border-color: var(--c-sand);   background: var(--c-nei-bg); }
      .callout-title {
        font-family: var(--font-ui);
        font-weight: 700;
        font-size: .85rem;
        letter-spacing: .06em;
        text-transform: uppercase;
        color: var(--c-forest);
        margin-bottom: .4rem;
      }
      .callout.warning .callout-title { color: var(--c-sienna); }
      .callout.neutral  .callout-title { color: var(--c-sand); }

      /* ── label chips ── */
      .chip {
        display: inline-block;
        font-family: var(--font-ui);
        font-size: .75rem;
        font-weight: 700;
        letter-spacing: .05em;
        padding: .15em .6em;
        border-radius: 3px;
      }
      .chip-sup  { background: var(--c-sup-bg); color: var(--c-moss);   }
      .chip-ref  { background: var(--c-ref-bg); color: var(--c-sienna); }
      .chip-nei  { background: var(--c-nei-bg); color: var(--c-sand);   }

      /* ── figures ── */
      figure { margin: 2rem 0; }
      figcaption { font-family: var(--font-ui); font-size: .8rem;
                    color: var(--c-muted); margin-top: .5rem; }
      img { max-width: 100%; border-radius: var(--r-card);
             box-shadow: 0 2px 8px rgba(0,0,0,.1); }

      /* ── code ── */
      pre, code {
        font-family: var(--font-mono);
        font-size: .8rem;
        background: #EAE8E3;
        border-radius: 4px;
      }
      pre { padding: 1rem; overflow-x: auto; }

      /* ── TOC ── */
      #quarto-sidebar, nav[role=doc-toc] {
        font-family: var(--font-ui);
        font-size: .82rem;
      }
      nav[role=doc-toc] a { color: var(--c-muted); text-decoration: none; }
      nav[role=doc-toc] a:hover { color: var(--c-moss); }
      nav[role=doc-toc] .active { color: var(--c-forest); font-weight: 600; }

      /* ── metric strip ── */
      .metrics {
        display: flex; gap: 1rem; flex-wrap: wrap; margin: 1.5rem 0;
      }
      .metric {
        flex: 1 1 140px;
        background: var(--c-card);
        border-radius: var(--r-card);
        padding: .9rem 1rem;
        box-shadow: 0 1px 4px rgba(0,0,0,.07);
        font-family: var(--font-ui);
      }
      .metric-value {
        font-size: 1.8rem; font-weight: 800; line-height: 1;
        color: var(--c-forest);
        font-variant-numeric: tabular-nums;
      }
      .metric-label {
        font-size: .75rem; color: var(--c-muted);
        text-transform: uppercase; letter-spacing: .06em;
        margin-top: .3rem;
      }

      /* ── assessment section ── */
      .assessment-block {
        background: var(--c-card);
        border-radius: var(--r-card);
        padding: 1.5rem;
        margin: 2rem 0;
        box-shadow: 0 1px 4px rgba(0,0,0,.07);
      }
      .assessment-block h4 {
        font-family: var(--font-ui);
        font-size: .8rem;
        text-transform: uppercase;
        letter-spacing: .07em;
        color: var(--c-muted);
        margin: 0 0 .5rem 0;
      }

      /* ── print ── */
      @media print {
        body { background: white; }
        .callout, .metric, .assessment-block { box-shadow: none; }
      }
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(arrow)
library(dplyr)
library(tidyr)
library(ggplot2)

DATASET <- "output/nli_scores/nli_config=deberta_zeroshot/assessment=GA1"

C_SUPPORTS <- "#2E6B4F"
C_REFUTES <- "#8B2E2E"
C_NEI <- "#A8896A"
LABEL_COLS <- c(
  SUPPORTS = C_SUPPORTS,
  NOT_ENOUGH_INFO = C_NEI,
  REFUTES = C_REFUTES
)

d <- open_dataset(DATASET) |>
  mutate(alignment = p_supports - p_refutes) |>
  collect()

n_total <- nrow(d)
n_bm <- n_distinct(d$bm)
n_km <- n_distinct(d$km)
pct_unc <- round(100 * mean(d$uncertain), 1)

label_overall <- d |>
  count(label) |>
  mutate(pct = round(100 * n / sum(n), 1))

pct_sup <- label_overall |> filter(label == "SUPPORTS") |> pull(pct)
pct_ref <- label_overall |> filter(label == "REFUTES") |> pull(pct)
pct_nei <- label_overall |> filter(label == "NOT_ENOUGH_INFO") |> pull(pct)
```

## Overview

This report summarises the first-pass NLI alignment scores for **GA1** (Global Assessment 1), produced by the `deberta_zeroshot` pipeline using `MoritzLaurer/deberta-v3-large-zeroshot-v2.0` at `max_length = 256`. Each citing work is scored against its partition's Background Message (BM) as
<span class="chip chip-sup">SUPPORTS</span>,
<span class="chip chip-ref">REFUTES</span>, or
<span class="chip chip-nei">NOT_ENOUGH_INFO</span>.

<div class="metrics">
<div class="metric">
  <div class="metric-value">`r format(n_total, big.mark=",")`</div>
  <div class="metric-label">Works scored</div>
</div>
<div class="metric">
  <div class="metric-value">`r n_bm`</div>
  <div class="metric-label">Background messages</div>
</div>
<div class="metric">
  <div class="metric-value">`r n_km`</div>
  <div class="metric-label">Key messages</div>
</div>
<div class="metric">
  <div class="metric-value">`r pct_unc`%</div>
  <div class="metric-label">Uncertain (conf < 0.60)</div>
</div>
</div>

<div class="callout warning">
<div class="callout-title">Interpretation caveat</div>
The 86 % uncertainty rate and near-chance median confidence (~0.47) reflect a known limitation of zero-shot NLI on this task, not a pipeline failure. The per-BM breakdown below shows the signal is highly structured — KM D performs well; KMs A–C show REFUTES inflation. Use the continuous alignment score (`p_supports − p_refutes`) for thresholding; do not act on raw argmax labels for KMs A–C without Phase-2 LLM review.
</div>

## Label distribution

### Overall

```{r label-overall}
label_overall |>
  mutate(
    label = factor(label, levels = c("SUPPORTS", "NOT_ENOUGH_INFO", "REFUTES"))
  ) |>
  arrange(label) |>
  knitr::kable(
    col.names = c("Label", "Works", "%"),
    format.args = list(big.mark = ",")
  )
```

<br>

```{r plot-overall, fig.height=2.2, fig.width=7, fig.cap="Overall label split — GA1, all 27 BMs combined."}
ggplot(
  label_overall,
  aes(
    x = pct,
    y = "GA1",
    fill = factor(label, levels = c("SUPPORTS", "NOT_ENOUGH_INFO", "REFUTES"))
  )
) +
  geom_col(width = 0.55) +
  geom_text(
    aes(label = paste0(pct, "%")),
    position = position_stack(vjust = 0.5),
    colour = "white",
    size = 3.5,
    fontface = "bold",
    family = "sans"
  ) +
  scale_fill_manual(
    values = LABEL_COLS,
    labels = c("SUPPORTS", "NOT_ENOUGH_INFO", "REFUTES")
  ) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(family = "sans", size = 10),
    plot.background = element_rect(fill = "#F5F3EE", colour = NA)
  )
```

### Per Key Message

```{r label-km}
d |>
  count(km, label) |>
  group_by(km) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  select(km, label, pct) |>
  pivot_wider(names_from = label, values_from = pct, values_fill = 0) |>
  select(km, SUPPORTS, NOT_ENOUGH_INFO, REFUTES) |>
  arrange(km) |>
  knitr::kable(col.names = c("KM", "SUPPORTS %", "NEI %", "REFUTES %"))
```

```{r plot-km, fig.height=3, fig.width=7, fig.cap="Label split per Key Message."}
d |>
  count(km, label) |>
  group_by(km) |>
  mutate(pct = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(
    label = factor(label, levels = c("SUPPORTS", "NOT_ENOUGH_INFO", "REFUTES"))
  ) |>
  ggplot(aes(x = pct, y = km, fill = label)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = paste0(round(pct), "%")),
    position = position_stack(vjust = 0.5),
    colour = "white",
    size = 3.2,
    fontface = "bold",
    family = "sans"
  ) +
  scale_fill_manual(values = LABEL_COLS) +
  labs(x = "% of citing works", y = NULL, fill = NULL) +
  theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    panel.grid.major.y = element_blank(),
    legend.position = "bottom",
    plot.background = element_rect(fill = "#F5F3EE", colour = NA)
  )
```

### Per Background Message

```{r plot-bm, fig.height=5, fig.width=10, fig.cap="Label split per BM, faceted by KM. D-series BMs show clearly stronger SUPPORTS signal."}
d |>
  count(km, bm, label) |>
  group_by(km, bm) |>
  mutate(pct = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(
    bm = factor(bm),
    label = factor(label, levels = c("SUPPORTS", "NOT_ENOUGH_INFO", "REFUTES"))
  ) |>
  ggplot(aes(x = bm, y = pct, fill = label)) +
  geom_col(width = 0.8) +
  facet_grid(. ~ km, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = LABEL_COLS) +
  labs(x = NULL, y = "% of citing works", fill = NULL) +
  theme_bw(base_size = 10, base_family = "sans") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.background = element_rect(fill = "#1C3A2E"),
    strip.text = element_text(colour = "white", face = "bold"),
    plot.background = element_rect(fill = "#F5F3EE", colour = NA),
    panel.background = element_rect(fill = "#FAFAF7")
  )
```

```{r table-bm}
d |>
  count(km, bm, label) |>
  group_by(km, bm) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  select(km, bm, label, pct) |>
  pivot_wider(names_from = label, values_from = pct, values_fill = 0) |>
  select(km, bm, SUPPORTS, NOT_ENOUGH_INFO, REFUTES) |>
  arrange(km, bm) |>
  knitr::kable(col.names = c("KM", "BM", "SUPPORTS %", "NEI %", "REFUTES %"))
```

## Confidence and alignment

### Confidence by label

```{r conf-table}
d |>
  group_by(label) |>
  summarise(
    n = n(),
    mean = round(mean(confidence, na.rm = TRUE), 3),
    median = round(median(confidence, na.rm = TRUE), 3),
    p25 = round(quantile(confidence, 0.25, na.rm = TRUE), 3),
    p75 = round(quantile(confidence, 0.75, na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  arrange(factor(
    label,
    levels = c("SUPPORTS", "NOT_ENOUGH_INFO", "REFUTES")
  )) |>
  knitr::kable(
    col.names = c("Label", "n", "Mean", "Median", "P25", "P75"),
    format.args = list(big.mark = ",")
  )
```

```{r plot-conf, fig.height=3.5, fig.width=7, fig.cap="Confidence (max class probability) density per label. The dashed line marks the uncertain threshold (0.60). Distributions barely separate — the model has limited certainty on this task."}
ggplot(d, aes(x = confidence, fill = label, colour = label)) +
  geom_density(alpha = 0.30, linewidth = 0.6) +
  geom_vline(
    xintercept = 0.60,
    linetype = "dashed",
    colour = "#3A4A41",
    linewidth = 0.5
  ) +
  annotate(
    "text",
    x = 0.62,
    y = Inf,
    label = "uncertain\nthreshold",
    hjust = 0,
    vjust = 1.4,
    size = 3,
    family = "sans",
    colour = "#3A4A41"
  ) +
  scale_fill_manual(values = LABEL_COLS) +
  scale_colour_manual(values = LABEL_COLS) +
  labs(x = "confidence", y = "density", fill = NULL, colour = NULL) +
  theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    legend.position = "bottom",
    plot.background = element_rect(fill = "#F5F3EE", colour = NA)
  )
```

### Alignment score per BM

The alignment score (`p_supports − p_refutes`) is a continuous summary on [−1, +1] that is more informative than the argmax label for downstream thresholding and ranking.

```{r align-table}
d |>
  group_by(km, bm) |>
  summarise(
    n = n(),
    mean_aln = round(mean(alignment, na.rm = TRUE), 3),
    sd_aln = round(sd(alignment, na.rm = TRUE), 3),
    pct_supp = round(100 * mean(label == "SUPPORTS"), 1),
    pct_ref = round(100 * mean(label == "REFUTES"), 1),
    pct_nei = round(100 * mean(label == "NOT_ENOUGH_INFO"), 1),
    pct_unc = round(100 * mean(uncertain), 1),
    .groups = "drop"
  ) |>
  arrange(km, bm) |>
  knitr::kable(
    col.names = c(
      "KM",
      "BM",
      "n",
      "Mean aln",
      "SD",
      "SUPP %",
      "REF %",
      "NEI %",
      "Unc %"
    ),
    format.args = list(big.mark = ",")
  )
```

```{r plot-aln, fig.height=3.5, fig.width=7, fig.cap="Distribution of alignment scores (p_supports − p_refutes) by label. Negative overall median (−0.065) reflects REFUTES inflation in KMs A–C; the KM D signal pulls the tail positive."}
ggplot(d, aes(x = alignment, fill = label, colour = label)) +
  geom_density(alpha = 0.30, linewidth = 0.6) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    colour = "#3A4A41",
    linewidth = 0.5
  ) +
  scale_fill_manual(values = LABEL_COLS) +
  scale_colour_manual(values = LABEL_COLS) +
  labs(
    x = "alignment score (p_supports − p_refutes)",
    y = "density",
    fill = NULL,
    colour = NULL,
    subtitle = "Positive → net support · Negative → net refutation"
  ) +
  theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    legend.position = "bottom",
    plot.background = element_rect(fill = "#F5F3EE", colour = NA)
  )
```

## Assessment

### KM D performs well; KMs A–C show systematic REFUTES inflation

<div class="assessment-block">
<h4>Key finding</h4>

**KM D is in a different class from A, B, C.** With 48.8 % SUPPORTS and only 14.6 % REFUTES overall, and individual BMs D1 (79.8 % SUPPORTS), D3 (78.6 %), and D12 (60.1 %) showing clear, consistent signals at comparatively lower uncertainty (65–84 %), KM D is the only part of GA1 where zero-shot NLI produces credible output. D3 and D1 have the best-calibrated scores in the entire dataset.

Likely reason: D-series BMs are probably more concrete and empirically testable — biodiversity metrics, species trends, ecosystem services — the kind of claims abstracts engage with directly.

</div>

<div class="callout warning">
<div class="callout-title">REFUTES inflation — KMs A, B, C</div>

Several BMs have REFUTES > SUPPORTS by a wide margin that is not scientifically credible:

| BM | SUPPORTS | REFUTES | Interpretation |
|----|---------|---------|----------------|
| B11 | 13 % | 56 % | Likely false positives |
| B5  | 27 % | 55 % | Likely false positives |
| C6  |  8 % | 55 % | Likely false positives |
| A3  |  4 % | 45 % | Likely false positives |
| A7  |  5 % | 43 % | Likely false positives |
| A8  | 15 % | 47 % | Likely false positives |

These are not credible. Citing papers of an assessment should not be 55 % refuting it. This is the well-documented zero-shot NLI failure mode: the hypothesis *"This paper refutes the following claim: …"* draws spurious entailment when the BM contains complex, multi-part, or politically nuanced language. The model fires on surface-level topical tension rather than genuine contradiction.

</div>

<div class="callout neutral">
<div class="callout-title">Noise BMs — effectively random</div>

Three BMs show SUPPORTS ≈ REFUTES — the argmax label is meaningless here:

- **A2** 29.5 % vs 29.9 % — coin flip
- **C2** 24.7 % vs 25.4 % — coin flip
- **B4** 29.0 % vs 35.4 % — near coin flip

For these, only the continuous alignment score and the raw probability triplet carry any information.

</div>

### Notable exception in KM A

**A5** (39.8 % SUPPORTS, 5.7 % REFUTES, mean alignment +0.095) is the only BM in KMs A–C with a clearly positive signal. Its claim is likely more concrete or empirically specific than its neighbours — worth inspecting manually as a calibration reference.

### Operational implications

1. **Do not act on raw REFUTES labels for KMs A–C** without Phase-2 LLM review. NLI here is a coarse NEI filter, not a judgment.
2. **KM D SUPPORTS at high confidence (> 0.70) can be treated as real** — D1, D3, D12 in particular.
3. **Use `p_supports − p_refutes` for ranking** rather than argmax labels. Papers with very high `p_nei` can be deprioritised for expert review.
4. **`max_length: 256` is probably hurting KMs A–C.** Truncating abstract tails removes findings — the part that matters most for support/refute judgements on conceptual BMs. The base-model-at-512 comparison (`deberta_zeroshot_v3`) is the highest-priority next experiment.
5. **Fine-tuning on BM citations** ([TD_NLI_training.md](TD_NLI_training.md)) would most help KMs A–C — the support/refute axis that zero-shot is fumbling. D-series BMs are already working.

---

*Pipeline: IPBES BM Fact Checker · NLI server: `ghcr.io/rkrug/nli-runpod` · Model read from `/health` at run time and recorded as `nli_model` in output parquet.*
