---
title: REFUTES Funnel
subtitle: Narrowing down citing works that contradict a Background Message
date: today
params:
  assessment_id: "GA1"
format:
    html:
        toc: true
        toc-depth: 4
        embed-resources: true
        code-fold: true
        code-summary: 'Show the code'
        keep-md: false
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)

# Self-contained-report helper, same idiom as IPBES_Fact_Checker.qmd's setup
# chunk -- but report_files_dir is built from params$assessment_id, NOT
# knitr::current_input(). This qmd is rendered once per assessment via
# quarto::quarto_render(output_file=, execute_params=); current_input()
# always returns the SOURCE filename ("IPBES_REFUTES_Report"), which would
# collide both assessment branches onto the same _files/ directory.
report_files_dir <- paste0("IPBES_REFUTES_Report_", params$assessment_id, "_files")
# Wipe before repopulating: copy_into_report_files() only copies in what it's
# given, so a file dropped from one run's inputs (e.g. a removed funnel
# level) would otherwise linger here forever across re-renders.
unlink(report_files_dir, recursive = TRUE, force = TRUE)
dir.create(report_files_dir, recursive = TRUE, showWarnings = FALSE)
copy_into_report_files <- function(paths) {
  file.copy(paths, report_files_dir, overwrite = TRUE)
  file.path(basename(report_files_dir), basename(paths))
}

data_path <- targets::tar_read(refutes_funnel_data) |>
  (\(x) x[grepl(paste0("_", params$assessment_id, "\\.rds$"), x)])()
x <- readRDS(data_path)

fig_paths <- copy_into_report_files(
  targets::tar_read(refutes_funnel_figures) |>
    (\(x) x[grepl(paste0("_", params$assessment_id, "\\.png$"), x)])()
)
fig_overall <- fig_paths[grepl("_overall_", fig_paths)]
fig_by_bm <- fig_paths[grepl("_by_bm_", fig_paths)]

table_paths <- copy_into_report_files(
  targets::tar_read(refutes_funnel_tables) |>
    (\(x) x[grepl(paste0("_", params$assessment_id, "\\.html$"), x)])()
)
table_l3 <- table_paths[grepl("_l3_", table_paths)]
```

# `r params$assessment_id`

## Methodology

Every work discovered by the snowball search under a Background Message
(BM) is put through a 3-level sieve, each level a **subset** of the one
before it, counted as **distinct citing works per (assessment, km, bm)** —
a work cited under two different BMs is two separate findings, not one, so
per-BM counts sum to the assessment-wide total shown below.

1. **Snowball corpus** — every citing work discovered for that BM (the
   forward/citing direction of the snowball search).
2. **NLI REFUTES (certain)** — of those, works with at least one claim
   scored `label == "REFUTES"` with `uncertain == FALSE` by the zero-shot
   NLI model (Phase 1 — see
   [TD_BM_NLI_approach](TD_BM_NLI_approach.html)).
3. **LLM-confirmed REFUTES** — of those, works where OpenRouter's LLM
   review (Phase 2 — see
   [TD_NLI_LLM_two_phase](TD_NLI_LLM_two_phase.html)) independently agreed
   with the REFUTES verdict (`llm_agrees == TRUE`).

There is deliberately no further "+ sufficient evidence" level: Phase 2's
own verification logic forces its `llm_label` to `NOT_ENOUGH_INFO` whenever
it judges the evidence insufficient, so an LLM verdict of REFUTES already
*implies* sufficient evidence — a separate level here would always contain
exactly the same works as level 3.

**Phase 2 (LLM verification) is a partial, ongoing process** — OpenRouter
credit has run out mid-run at least once already, so level 3 below
reflects only the pairs reviewed so far, not the full REFUTES-certain set
from level 2. This number will grow as `llm_verification_parquet` gets
more OpenRouter runs; re-render this report after a rerun rather than
treating today's snapshot as final.

```{r}
#| label: empty-check
#| results: asis
if (isTRUE(x$empty)) {
  cat("No data available yet for this assessment — the snowball, NLI, or LLM verification stage hasn't produced output here.\n\n")
  knitr::knit_exit()
}
```

## Overall numbers

```{r}
#| label: overall-table
knitr::kable(
  x$funnel_overall,
  col.names = c("Level", "Description", "Distinct works")
)
```

```{r}
#| label: overall-figure
#| results: asis
if (length(fig_overall)) cat(sprintf("![](%s)\n\n", fig_overall))
```

## Per-BM numbers

```{r}
#| label: by-bm-table
knitr::kable(
  x$funnel_by_bm,
  col.names = c(
    "KM", "BM", "Corpus", "NLI REFUTES", "LLM-confirmed",
    "% 2 of 1", "% 3 of 2"
  )
)
```

```{r}
#| label: by-bm-figure
#| results: asis
if (length(fig_by_bm)) cat(sprintf("![](%s)\n\n", fig_by_bm))
```

## Level 3: LLM-confirmed REFUTES

Every citing work where the LLM's own independent review agreed with NLI's
REFUTES verdict (and, by construction, judged the evidence for it
sufficient — see Methodology). Use the BM dropdown in the filter row below
the header to narrow to one Background Message.

```{r}
#| label: level3-table
#| results: asis
if (length(table_l3)) {
  cat(sprintf(
    '<iframe src="%s" width="100%%" height="800px" style="border:none;"></iframe>\n\n',
    table_l3
  ))
}
```
