---
title: Fact Checker DRAFT
subtitle: Fact Checker of BMs of IPBES Assessments
date: today
author:
  - name: Rainer M Krug 
    id: rmk
    orcid: 1234-1234-1234-1234
    email: Rainer@krugs.de
    affiliation: 
      - name: University of Zürich
        city: Zürich
        state: ZH
        url: www.uzh.ch
    roles: [author, editor]
abstract: > 
  The assessments Linked Open Data is used to conduct analysis and searches 
  to discover new knowledge related to the assessmwents.
# keywords:
#   - aaaaa
#   - bbbbb
license: "CC BY"
citation: 
  type: report
  doi: xxxxxxxxxxxxxx
doi: xxxxxxxxxxxxxx4
version: 0.0.1

format:
    html:
        toc: true
        toc-depth: 6
        toc_expand: true
        embed-resources: false
        code-fold: true
        code-summary: 'Show the code'
        keep-md: false
        # grid:
        #     sidebar-width: 0px
        #     body-width: 4000px
        #     margin-width: 200px
        #     gutter-width: 1.5rem    
---

```{r}
#| label: setup
#| include: false

library(openalexR)
library(dplyr)
library(arrow)
library(ggplot2)

library(IPBES.R)
library(htmlwidgets)
library(htmltools)

library(tictoc)
```

# Introduction
IPBES assessments were converted / enhanced into [Linked Open Data (LOD)](https://github.com/IPBES-Data/IPBES_LOD) using the ttl format. 

Three assessments available in ttl format are currently:

- the [First Global Assessment](https://github.com/IPBES-Data/IPBES_LOD/blob/main/Global%20Assessment%201/README_GA1.md) (GA1)
- the [Invasive Alien Species Assessment](https://github.com/IPBES-Data/IPBES_LOD/tree/main/Invasive%20Alien%20Species%20Assessment) (IAS).
<!-- - the [Values Assessment](https://github.com/IPBES-Data/IPBES_LOD/tree/main/Values%20Assessment) (VA) -->

# Methods

Methodology is documented separately in the `TD_*.md` design documents at the
repository root. **Note: these are currently stale relative to the
implementation** — treat them as a record of the design rationale rather than
an up-to-date specification of the pipeline as it stands today.

- [TD_targets.md](TD_targets.md) — the `targets` pipeline architecture
- [TD_BM_NLI_approach.md](TD_BM_NLI_approach.md) — the active NLI scoring approach
- [TD_LLM_approach.md](TD_LLM_approach.md) — the parked LLM-comparison approach
- [TD_NLI_LLM_two_phase.md](TD_NLI_LLM_two_phase.md) — the two-phase NLI → LLM design
- [TD_NLI_training.md](TD_NLI_training.md) — NLI fine-tuning plan

The figures and tables in the Results section below are all produced by
`targets` pipeline targets — see `_targets.R` and the corresponding
`R/build_*.R` files for the exact computation behind each one.

# Results

```{r}
#| label: results_data
#| include: false

refs_paths <- targets::tar_read(refs_parquet)
works_paths <- targets::tar_read(works_parquet)
works_citing_paths <- targets::tar_read(works_citing_parquet)

refs <- dplyr::bind_rows(lapply(refs_paths, function(p) {
  arrow::open_dataset(p) |> dplyr::select(doi) |> dplyr::collect()
}))

key_papers <- dplyr::bind_rows(lapply(works_paths, function(p) {
  arrow::open_dataset(p) |>
    dplyr::select(km, bm, doi, publication_year, type) |>
    dplyr::collect()
})) |>
  dplyr::mutate(oa_input = TRUE)

citing_papers <- dplyr::bind_rows(lapply(works_citing_paths, function(p) {
  arrow::open_dataset(p) |>
    dplyr::select(km, bm, doi, publication_year, type) |>
    dplyr::collect()
})) |>
  dplyr::mutate(oa_input = FALSE)

all_papers <- dplyr::bind_rows(key_papers, citing_papers)

# Figure and table artifact paths — resolved via tar_read() so the report
# always shows what the pipeline actually last built, never a stale/hardcoded
# path.
fig_pub_per_year_paths <- targets::tar_read(fig_pub_per_year)
fig_pub_per_year_png <- fig_pub_per_year_paths[grepl(
  "\\.png$",
  fig_pub_per_year_paths
)]
fig_pub_per_year_pdf <- fig_pub_per_year_paths[grepl(
  "\\.pdf$",
  fig_pub_per_year_paths
)]
fig_pub_per_year_svg <- fig_pub_per_year_paths[grepl(
  "\\.svg$",
  fig_pub_per_year_paths
)]

overlap_key_paper_html <- targets::tar_read(overlap_key_paper_table)
overlap_key_paper_html <- overlap_key_paper_html[grepl(
  "\\.html$",
  overlap_key_paper_html
)]

overlap_after_2018_sub_messages_html <- targets::tar_read(
  overlap_after_2018_sub_messages_table
)
overlap_after_2018_sub_messages_html <- overlap_after_2018_sub_messages_html[grepl(
  "\\.html$",
  overlap_after_2018_sub_messages_html
)]

overlap_after_2018_background_messages_html <- targets::tar_read(
  overlap_after_2018_background_messages_table
)
overlap_after_2018_background_messages_html <- overlap_after_2018_background_messages_html[grepl(
  "\\.html$",
  overlap_after_2018_background_messages_html
)]
```

The dataset contains **`r refs |> dplyr::distinct(doi) |> nrow()`** papers
referenced in the background messages. Out of these, **`r key_papers |> dplyr::distinct(doi) |> nrow()`**
were usable for the snowball search and also yielded results.

```{r}
#| label: numbers_table
#|

all_papers |>
  dplyr::summarise(
    n = dplyr::n(),
    n_key_paper = sum(oa_input, na.rm = TRUE),
    n_non_key_paper = sum(!oa_input, na.rm = TRUE),
    n_nkp_pre_2018 = sum(
      (!oa_input) & (publication_year <= 2018),
      na.rm = TRUE
    ), # Count of citing papers with publication_year <= 2018
    n_nkp_post_2018 = sum(
      (!oa_input) & (publication_year > 2018),
      na.rm = TRUE
    ), # Count of citing papers with publication_year > 2018
    .by = c(km, bm)
  ) |>
  dplyr::arrange(km, bm) |>
  IPBES.R::table_dt()
```

## Publication types after 2018

Here we show the different types of publication after 2018 as identified by the snowbal search. All Sub-background messages are grouped.

```{r}
#| label: publication_types_grouped
#| echo: false

all_papers |>
  dplyr::filter(
    publication_year > 2018
  ) |>
  dplyr::summarize(
    n = dplyr::n(),
    .by = type
  ) |>
  dplyr::arrange(
    dplyr::desc(n)
  ) |>
  knitr::kable()
```

This table shows the number of work types per sub-background message. It allows filtering as well as sorting and downloading.

```{r}
#| label: publication_types_non_grouped
#| echo: false

all_papers |>
  dplyr::filter(
    publication_year > 2018
  ) |>
  dplyr::summarize(
    n = dplyr::n(),
    .by = c(type, km, bm)
  ) |>
  dplyr::arrange(
    dplyr::desc(n)
  ) |>
  IPBES.R::table_dt(
    fixed_columns = NULL
  )
```

## Publications per year

![](`r fig_pub_per_year_png`)

- [Download `png`](`r fig_pub_per_year_png`)
- [Download `pdf`](`r fig_pub_per_year_pdf`)
- [Download `svg`](`r fig_pub_per_year_svg`)

## Overlap

In this section we will look at the number of background messages in which a given paper, as identified by their OpenAlex id. We willlook at that separatr for key-papers and non key-papers.

In the following tables only the papers which occur in more than one background message are included and only papers which were published after 2018.

This overlap can have different causes. Either there is an overlap of the key-papers, ehich results in an overlap of the new publications, or there is a topical overlap betweent the new paper themselves. 

### Overlap of Key Papers in Background Messages

The first table shows the overlap of the key-papers only regarding background messages.

Only papers occurring in more than 1 background message are shown.

<iframe src="`r overlap_key_paper_html`" width="100%" height="500px"></iframe>

### Overlap of Papers after 2018 in Sub Messages

The second table shows the overlap of all papers after 2018 regarding sub messages.

Only papers with in more then 5 Sub Messages are shown.

<iframe src="`r overlap_after_2018_sub_messages_html`" width="100%" height="500px"></iframe>

### Overlap of Papers after 2018 in Background Messages

The third table shows the overlap of all papers after 2018 regarding background messages.

Only papers with in more then 5 Background Messages are shown.

<iframe src="`r overlap_after_2018_background_messages_html`" width="100%" height="500px"></iframe>

## NLI Alignment Scores

Each citing work is scored against its partition's Background Message (BM) as `SUPPORTS`, `REFUTES`, or `NOT_ENOUGH_INFO` using a zero-shot NLI model. See [TD_BM_NLI_approach.md](TD_BM_NLI_approach.md) for methodology.

```{r}
#| label: nli_overview_paths
#| include: false

nli_overview_data_paths <- targets::tar_read(nli_overview_data)
nli_overview_figures_paths <- targets::tar_read(nli_overview_figures)
```

```{r}
#| label: nli_overview_sections
#| results: asis

for (data_path in nli_overview_data_paths) {
  x <- readRDS(data_path)
  cat(sprintf("\n### %s\n\n", x$assessment))

  if (isTRUE(x$empty)) {
    cat("No NLI scores available yet for this assessment.\n\n")
    next
  }

  cat(sprintf(
    "- **%s** works scored\n- **%d** background messages\n- **%d** key messages\n- **%s%%** uncertain (confidence < 0.60)\n\n",
    format(x$n_total, big.mark = ","),
    x$n_bm,
    x$n_km,
    x$pct_unc
  ))

  fig_this <- nli_overview_figures_paths[
    grepl(paste0("_", x$assessment, "\\.png$"), nli_overview_figures_paths)
  ]
  plot_overall <- fig_this[grepl("plot_overall", fig_this)]
  plot_km <- fig_this[grepl("plot_km", fig_this)]
  plot_bm <- fig_this[grepl("plot_bm", fig_this)]
  plot_conf <- fig_this[grepl("plot_conf", fig_this)]
  plot_aln <- fig_this[grepl("plot_aln", fig_this)]

  cat("#### Label distribution\n\n")
  cat(
    knitr::kable(
      x$label_overall |>
        dplyr::mutate(
          label = factor(
            label,
            levels = c("SUPPORTS", "NOT_ENOUGH_INFO", "REFUTES")
          )
        ) |>
        dplyr::arrange(label),
      col.names = c("Label", "Works", "%")
    ),
    sep = "\n"
  )
  cat("\n\n")
  if (length(plot_overall)) {
    cat(sprintf("![](%s)\n\n", plot_overall))
  }

  cat("#### Per Key Message\n\n")
  cat(
    knitr::kable(
      x$label_km,
      col.names = c("KM", "SUPPORTS %", "NEI %", "REFUTES %")
    ),
    sep = "\n"
  )
  cat("\n\n")
  if (length(plot_km)) {
    cat(sprintf("![](%s)\n\n", plot_km))
  }

  cat("#### Per Background Message\n\n")
  if (length(plot_bm)) {
    cat(sprintf("![](%s)\n\n", plot_bm))
  }
  cat(
    knitr::kable(
      x$table_bm,
      col.names = c("KM", "BM", "SUPPORTS %", "NEI %", "REFUTES %")
    ),
    sep = "\n"
  )
  cat("\n\n")

  cat("#### Confidence by label\n\n")
  cat(
    knitr::kable(
      x$conf_table,
      col.names = c("Label", "n", "Mean", "Median", "P25", "P75")
    ),
    sep = "\n"
  )
  cat("\n\n")
  if (length(plot_conf)) {
    cat(sprintf("![](%s)\n\n", plot_conf))
  }

  cat("#### Alignment score per BM\n\n")
  cat(
    knitr::kable(
      x$align_table,
      col.names = c(
        "KM",
        "BM",
        "n",
        "Mean aln",
        "SD",
        "SUPP %",
        "REF %",
        "NEI %",
        "Unc %"
      )
    ),
    sep = "\n"
  )
  cat("\n\n")
  if (length(plot_aln)) cat(sprintf("![](%s)\n\n", plot_aln))
}
```

### GA1 — Detailed Interpretation (static, 2026-07-01)

::: {.callout-note}
## About this section
Written against an earlier sample of GA1 scores and does **not** auto-update from the live data above — treat as illustrative commentary pending a refreshed pass once scoring is more complete.
:::

**Key finding.** **KM D is in a different class from A, B, C.** With 48.8 % SUPPORTS and only 14.6 % REFUTES overall, and individual BMs D1 (79.8 % SUPPORTS), D3 (78.6 %), and D12 (60.1 %) showing clear, consistent signals at comparatively lower uncertainty (65–84 %), KM D is the only part of GA1 where zero-shot NLI produces credible output. D3 and D1 have the best-calibrated scores in the entire dataset.

Likely reason: D-series BMs are probably more concrete and empirically testable — biodiversity metrics, species trends, ecosystem services — the kind of claims abstracts engage with directly.

::: {.callout-warning}
## REFUTES inflation — KMs A, B, C

Several BMs have REFUTES > SUPPORTS by a wide margin that is not scientifically credible:

| BM | SUPPORTS | REFUTES | Interpretation |
|----|---------|---------|----------------|
| B11 | 13 % | 56 % | Likely false positives |
| B5  | 27 % | 55 % | Likely false positives |
| C6  |  8 % | 55 % | Likely false positives |
| A3  |  4 % | 45 % | Likely false positives |
| A7  |  5 % | 43 % | Likely false positives |
| A8  | 15 % | 47 % | Likely false positives |

These are not credible. Citing papers of an assessment should not be 55 % refuting it. This is the well-documented zero-shot NLI failure mode: the hypothesis *"This paper refutes the following claim: …"* draws spurious entailment when the BM contains complex, multi-part, or politically nuanced language. The model fires on surface-level topical tension rather than genuine contradiction.
:::

::: {.callout-note}
## Noise BMs — effectively random

Three BMs show SUPPORTS ≈ REFUTES — the argmax label is meaningless here:

- **A2** 29.5 % vs 29.9 % — coin flip
- **C2** 24.7 % vs 25.4 % — coin flip
- **B4** 29.0 % vs 35.4 % — near coin flip

For these, only the continuous alignment score and the raw probability triplet carry any information.
:::

#### Notable exception in KM A

**A5** (39.8 % SUPPORTS, 5.7 % REFUTES, mean alignment +0.095) is the only BM in KMs A–C with a clearly positive signal. Its claim is likely more concrete or empirically specific than its neighbours — worth inspecting manually as a calibration reference.

#### Operational implications

1. **Do not act on raw REFUTES labels for KMs A–C** without Phase-2 LLM review. NLI here is a coarse NEI filter, not a judgment.
2. **KM D SUPPORTS at high confidence (> 0.70) can be treated as real** — D1, D3, D12 in particular.
3. **Use `p_supports − p_refutes` for ranking** rather than argmax labels. Papers with very high `p_nei` can be deprioritised for expert review.
4. **`max_length: 256` is probably hurting KMs A–C.** Truncating abstract tails removes findings — the part that matters most for support/refute judgements on conceptual BMs. The base-model-at-512 comparison (`deberta_zeroshot_v3`) is the highest-priority next experiment.
5. **Fine-tuning on BM citations** ([TD_NLI_training.md](TD_NLI_training.md)) would most help KMs A–C — the support/refute axis that zero-shot is fumbling. D-series BMs are already working.


