---
title: "QA: NLI Model Benchmark"
subtitle: Held-out evaluation across the zero-shot baseline and every fine-tuned checkpoint
date: today
format:
    html:
        toc-depth: 5
        toc_expand: true
---

```{r}
#| label: targets-dependencies
#| include: false

# Named literally so tar_quarto() turns this into a real dependency edge.
# Hand-written rather than generated from config.yaml's `reports:` block:
# there is no dimension to expand over -- one report covers every model.
qa <- readRDS(targets::tar_read(nli_benchmark_qa_data))

library(dplyr)
library(ggplot2)
library(knitr)

pct <- function(x) ifelse(is.na(x), "--", sprintf("%.1f%%", 100 * x))
```

## What this measures, and what it does not

**It measures agreement with `gpt-4o-mini`.** Every label in the holdout comes
from Phase 2's grounded, quote-checked LLM verdict. Fine-tuning distils that
judgment into the NLI model and this report scores how well the distillation
worked. Every table below is therefore reported **twice**:

- `reference = "llm"` — truth is gpt-4o-mini's Phase 2 verdict. This measures
  *fidelity to the judge*, not correctness, and is what earlier versions of this
  report showed on their own. A model that perfectly reproduced a wrong judge
  would score perfectly here.
- `reference = "human"` — truth is the adjudicated verdict of two independent
  reviewers. **This is the one to quote.**

Under the human reference gpt-4o-mini stops being the yardstick and becomes just
another predictor, scored on the same rows as every NLI model. The gap between
the two references is the quantity this report exists to expose.

Design and reasoning: [TD_NLI_training](TD_NLI_training.html).

```{r}
#| label: fold
#| echo: false
f <- qa$fold
kable(data.frame(
  Property = c("benchmark salt", "holdout fraction", "monitor fraction", "grouping key",
               "train rows", "monitor rows", "holdout rows",
               "train groups", "monitor groups", "holdout groups",
               "claims in ANY two folds", "works in more than one fold",
               "holdout rows involving a shared work",
               "holdout rows with a human label"),
  Value = c(f$salt, f$holdout_fraction, f$monitor_fraction, f$group_key,
            f$n_train, f$n_monitor, f$n_holdout,
            f$n_groups_train, f$n_groups_monitor, f$n_groups_holdout,
            f$claims_in_both_folds,
            sprintf("%d of %d (%s)", f$works_in_both_folds, f$works_total,
                    pct(f$works_in_both_folds / f$works_total)),
            sprintf("%d (%s)", f$holdout_rows_with_shared_work,
                    pct(f$holdout_rows_with_shared_work / f$n_holdout)),
            sprintf("%d of %d sampled", f$n_gold_matched, f$n_gold))
), caption = "Fold composition and leakage diagnostics")
```

**There are three folds, not two.** `monitor` is what fine-tuning evaluates on
and selects its best checkpoint from; `holdout` is scored here and read by
nothing during training. They were previously the same fold, so
`load_best_model_at_end` picked the checkpoint that scored best on the very rows
this report then published — every number from before that change is optimistic
by an unmeasured margin.

**`claims in ANY two folds` must be 0.** The holdout is grouped by
`(assessment, bm)` — deliberately *without* `km`, because one Background
Message belongs to several Key Messages and carries byte-identical claim text
under each. Grouping on `(assessment, km, bm)` looked tighter and was measurably
leakier: 141 groups, 126 claim strings present in both folds. If this number is
ever non-zero, the grouping key has drifted.

**Work overlap is reported, not eliminated.** A work cited under several BMs can
appear on both sides paired with *different* claims. That is much weaker than
claim-text leakage — the model sees the premise, never the pair or its label —
and closing it needs a connected-component split that at this size yields a few
huge unbalanced components, trading a small measurable bias for a large
unmeasurable one. Judge the headline numbers with the percentage above in mind.

## Models under test

```{r}
#| label: models
#| echo: false
kable(qa$models |> select(model_id, mode, split_aware, max_length, n_rows, n_truncated),
      caption = "Scored models. `zeroshot` = 3 forward passes per row with the deployed hypothesis template, entailment logits cross-normalised. `direct` = the fine-tuned head's own 3-way softmax. `split_aware` is NA for the untrained baseline.")
```

::: {.callout-important}
## `split_aware = FALSE` means the scores below are contaminated

A checkpoint trained **before** the holdout existed was trained on rows drawn
from the whole pool — including what is now the test fold. Its score here is
recall on memorised rows, not generalisation.

This is not hypothetical. The larger pre-split run trained on 558 of the
corpus's 562 REFUTES rows, so roughly **169 of the test fold's 170 REFUTES were
in its training data**, and it duly posts the best numbers in every table below.

Those runs are shown rather than hidden, because they are the clearest available
evidence for *how badly a leaking split flatters a model* — which is the finding
that motivated this benchmark. But **no table below is a ranking**. Compare only
rows where `split_aware` is `TRUE` against the baseline.
:::

One asymmetry that runs the other way, and makes the honest result
conservative: the fine-tuned models are scored at the 512 tokens they were
trained on, and 141 of 467 premises (30%) hit that limit, against 3 for the
baseline at its deployed 2048. The fine-tuned models are working from truncated
abstracts where the baseline is not.

The baseline is scored the way it is actually **deployed** — a binary
entailment head cannot emit SUPPORTS/REFUTES/NEI directly, so each candidate
label is folded into its own reformulated hypothesis and the three entailment
logits are cross-normalised. A baseline scored differently from how it runs in
production is not a baseline.

One caveat on exactness: the deployed pipeline cleans the premise
(`clean_title()`/`clean_abstract()` strip HTML/JATS/LaTeX) while this benchmark
uses raw `title`/`abstract`, matching what `train_nli.py` trains on. Spot-checked
against real deployed scores, local and production probabilities agree to ~4
decimal places on clean abstracts and diverge on marked-up ones.

## The human ceiling

Two reviewers labelled the same sample independently, blind to every model's
output. How often they agreed with *each other* bounds how well any automated
system can be expected to agree with the gold standard — a model scoring above
this line is matching the adjudicated subset, not outperforming human judgement.

Cohen's κ rather than raw agreement: around 95% of this corpus is
`NOT_ENOUGH_INFO`, so two reviewers who both default to it agree over 90% of the
time while carrying no information at all. κ prices that baseline in.

```{r}
#| label: agreement
#| echo: false
qa$agreement |>
  transmute(pair, n, agreement = pct(agreement), kappa = round(kappa, 3)) |>
  kable(caption = "Pairwise agreement. The first row is the ceiling; everything below is read against it.")
```

Rows the two reviewers could not agree on are excluded from scoring — they are
reported as the ceiling, not resolved by a third opinion. The rule was fixed
before any reviewing started, so it cannot have been chosen to flatter a result.

## The headline: SUPPORTS → REFUTES recovery

60% of every confirmed refutation in this corpus is a pair the zero-shot model
called **SUPPORTS** and the LLM overturned. It did not merely miss them — it got
them backwards. So the question that decides whether fine-tuning did this
project's job is: of the holdout rows where NLI said SUPPORTS and the truth is
REFUTES, how many does each model now call REFUTES?

```{r}
#| label: flip
#| echo: false
qa$flip |>
  left_join(qa$models |> select(model, model_id, split_aware), by = "model") |>
  mutate(model_id = coalesce(model_id, model)) |>
  transmute(reference, model_id, split_aware, n_subset, n_recovered,
            recovery_rate = pct(recovery_rate)) |>
  arrange(desc(reference), desc(n_recovered)) |>
  kable(caption = "Rows where NLI said SUPPORTS and the truth is REFUTES. Read only the split_aware = TRUE rows against the baseline (NA).")
```

A checkpoint that does not move this number has not earned deployment, whatever
its macro-F1 says.

## Per-class performance

Accuracy is deliberately absent. The holdout is near-balanced while deployment
is 0.2–3% REFUTES, so a balanced-set accuracy flatters every model equally and
predicts nothing about the real stream. REFUTES precision and recall are given
separately and never folded into F1 for the headline: a missed refutation is a
finding that never surfaces, a false one is a claim a human discards in seconds,
and F1 assumes those cost the same.

```{r}
#| label: per-class
#| echo: false
qa$per_class |>
  left_join(qa$models |> select(model, model_id, split_aware), by = "model") |>
  mutate(model_id = coalesce(model_id, model)) |>
  transmute(reference, model_id, split_aware, label,
            precision = pct(precision), recall = pct(recall),
            f1 = sprintf("%s [%s, %s]", pct(f1), pct(f1_lo), pct(f1_hi)),
            support) |>
  arrange(desc(reference), label, model_id) |>
  kable(caption = "F1 is shown with its 95% bootstrap interval. Two models whose intervals overlap are not distinguishable on this holdout.")
```

```{r}
#| label: confusion
#| echo: false
qa$confusion |>
  left_join(qa$models |> select(model, model_id), by = "model") |>
  mutate(model_id = coalesce(model_id, model)) |>
  tidyr::pivot_wider(id_cols = c(reference, model_id, truth), names_from = pred_label,
                     values_from = n, values_fill = 0) |>
  arrange(desc(reference), model_id, truth) |>
  kable(caption = "Confusion matrix: rows = truth (the LLM verdict, or the adjudicated human verdict), columns = model prediction")
```

### Weighted rates — these are the corpus-level numbers

The human sample deliberately over-draws REFUTES, so **the unweighted table
above describes the sample, not the corpus**. Reweighting by inverse inclusion
probability recovers corpus-level rates.

**Both precision and recall need it.** An earlier version of this report
reweighted only precision, on the reasoning that recall is unbiased under
stratification. That holds when the stratification is on the *true* label. Here
it is on the **LLM's label** — the thing under question — and every NLI model
scored was distilled from that same LLM, so its predictions correlate with the
stratification variable and the sample is skewed *within* each truth class.
Measured on the GA1 sample: gpt-4o-mini's SUPPORTS recall is 40.4% unweighted
against 75.4% weighted. Between-model *ranking* on identical rows is unaffected.

```{r}
#| label: weighted
#| echo: false
if (is.null(qa$per_class_weighted)) {
  cat("_No sample manifest found — weighted rates unavailable._")
} else {
  qa$per_class_weighted |>
    left_join(qa$models |> select(model, model_id), by = "model") |>
    mutate(model_id = coalesce(model_id, model)) |>
    transmute(model_id, label,
              precision = pct(precision_weighted), recall = pct(recall_weighted),
              f1 = pct(f1_weighted),
              est_share = pct(support_weighted / sum(support_weighted))) |>
    arrange(label, model_id) |>
    kable(caption = "Corpus-level rates, inverse-inclusion-probability weighted (human reference). `est_share` is that class's estimated share of the holdout — a tiny share means the sample barely saw the class, so read its rates as indicative only.")
}
```

::: {.callout-warning}
## REFUTES is not estimable from this sample

The sample drew 100 of 200 rows from the REFUTES stratum and yielded **one**
adjudicated REFUTES, because the judge's REFUTES precision is a few per cent.
Stratifying on a label that is mostly wrong is not an enrichment.

So the SUPPORTS → REFUTES recovery rate above, and every REFUTES per-class
figure, rests on a support of ~1 and must not be read as a measurement. What
this sample *does* establish — and it is the question that motivated building a
gold standard at all — is that the judge's REFUTES calls are overwhelmingly not
refutations.
:::

## Threshold sweep

Phase 1 stores the full probability distribution and every threshold decision is
made downstream in `dplyr`, so a single argmax operating point would describe
neither the model nor the deployment. Framed as a REFUTES detector — predict
REFUTES iff `p_refutes >= t` — which is what these scores are actually used for.
This is also where the re-calibrated deployment threshold comes from: the
base-rate shift between a balanced training fold and a 0.2% deployment stream
guarantees an argmax classifier over-predicts REFUTES in production.

```{r}
#| label: sweep
#| echo: false
#| fig-width: 9
#| fig-height: 5
qa$threshold_sweep |>
  left_join(qa$models |> select(model, model_id), by = "model") |>
  tidyr::pivot_longer(c(precision, recall), names_to = "metric", values_to = "value") |>
  ggplot(aes(threshold, value, colour = model_id, linetype = metric)) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "p_refutes threshold", y = NULL, colour = "model", linetype = NULL,
       title = "REFUTES detection: precision and recall across thresholds") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box = "vertical")
```

## Per assessment

558 of 562 REFUTES in the pool come from GA1 and IAS, so a pooled number mostly
measures those two. TCA and VA are shown separately even though their counts are
tiny — "we cannot tell" is a legitimate and useful result, and whether anything
learned here generalises beyond GA1/IAS is an open question.

```{r}
#| label: by-assessment
#| echo: false
qa$by_assessment |>
  left_join(qa$models |> select(model, model_id), by = "model") |>
  transmute(model_id, assessment, n, n_refutes,
            refutes_recall = pct(refutes_recall),
            refutes_precision = pct(refutes_precision)) |>
  arrange(assessment, model_id) |>
  kable()
```

## What this cannot tell you

**Recall against the real corpus.** The holdout inherits Phase 2's routing bias:
for citing works only `REFUTES-certain` and `SUPPORTS-certain` pairs were ever
sent for LLM review, so refutations hiding among the 1.23M `NOT_ENOUGH_INFO`
pairs in GA1 alone are in no fold of any split. The full-coverage key-paper chain
estimates 0.37% of NEI pairs are genuine refutations against 3.09% of REFUTES
pairs — an 8x enrichment across a 4.8x larger pile, meaning **37% of the
refutations found there were sitting in the NEI bucket**. Every number above is
conditional on a pair having been routed for review in the first place.

Closing that needs a recall probe: a random sample across the full NLI output
including NEI, labelled by Phase 2. See [TD_NLI_training](TD_NLI_training.html).
