<!-- Shared BODY for the QA_BM_Split_Report report family.
     Rendered only via the generated wrapper .qmd files that `{{< include >}}`
     this file -- see R/generate_report_wrappers.R and config.yaml's `reports:`
     section. The leading underscore keeps Quarto from rendering it on its own
     -- but it also keeps it out of the project entirely, so tar_quarto() never
     scans this file for dependencies. That is why the tar_read() calls live in
     the wrapper and arrive here as `report_deps`; `params$...` likewise comes
     from the wrapper. -->

```{r}
#| label: setup
#| include: false

library(dplyr)

# No report_files_dir/copy_into_report_files() here -- unlike the earlier
# DT::datatable() version of this report, the highlighted-text view is
# plain HTML embedded directly in this document (no separate widget file
# to ship as a sidecar), so there's nothing to copy in and no iframe/
# resource-path concerns at all.
gran_suffix <- paste0("_", params$granularity)

expected_name <- function(ext) {
  sprintf(
    "bm_split_highlighted_%s%s.%s",
    params$assessment_id,
    gran_suffix,
    ext
  )
}

data_path <- report_deps$bm_split_report_highlighted |>
  (\(x) x[basename(x) == expected_name("rds")])()
# config.yaml's `reports:` declares every (assessment x granularity)
# combination, but a granularity that has not been built yet produces no rds
# at all -- different from an rds that exists and reports x$empty, and it used
# to abort the WHOLE project render on the first such combination rather than
# just that one document. Route both into the same empty-state below.
data_path <- data_path[file.exists(data_path)]
x <- if (length(data_path) == 1L) readRDS(data_path) else list(empty = TRUE)

if (isTRUE(x$empty)) {
  n_bm <- 0L; n_claims <- 0L; has_confidence <- FALSE; pct_with_confidence <- NA
} else {
n_bm <- x$data |> distinct(km, bm) |> nrow()
n_claims <- nrow(x$data)
has_confidence <- "confidence" %in% names(x$data) && any(!is.na(x$data$confidence))
pct_with_confidence <- if (has_confidence) {
  round(100 * mean(!is.na(x$data$confidence)), 1)
} else {
  NA
}
}
```

```{r}
#| label: empty-check
#| results: asis
# This body had no empty-state at all, unlike its five siblings -- a missing
# input was an error rather than a rendered explanation.
if (isTRUE(x$empty)) {
  cat(sprintf(
    "No BM-split QA data for `%s` at granularity `%s` -- that granularity has not been segmented/scored yet. Run the `nli_ready_evidence_parquet` chain for it and re-render.\n\n",
    params$assessment_id, params$granularity
  ))
  knitr::knit_exit()
}
```

# `r params$assessment_id` -- granularity: `r params$granularity`

## Methodology

Every Background Message (BM) in this assessment is split into claims by
`nli_ready_evidence_parquet` under the `granularity` of the NLI config
selected by `input/config.yaml`'s `fact_checking.nli`. This report is a
**QA check on that split, not a scoring result** — for every BM field, it
shows the original `bm_description`/`bm_label` text with each resulting
claim's source span colour-highlighted, followed by the claims themselves
as a colour-matched list, so the segmentation can be sanity-checked by eye
before any NLI/LLM scoring is run against it. Highlighting is based on the
raw span each claim was cut from, before any confidence-qualifier removal
or (for `atomic_bm`) LLM completion -- so it covers at least the major
part of a claim's origin, even where the listed claim text itself was
reworded. See `TD_BM_NLI_approach.qmd` for the full design rationale
behind each granularity option (`naive_bm` / `complete_bm` / `atomic_bm`).

```{r}
#| label: summary
#| results: asis
cat(sprintf("- **%d** distinct BMs, **%d** resulting claims\n", n_bm, n_claims))
if (has_confidence) {
  cat(sprintf(
    "- **%.1f%%** of claims had an IPBES confidence qualifier (\"well established\" / \"established but incomplete\" / \"unresolved\" / \"inconclusive\") extracted out of the claim text into its own column\n",
    pct_with_confidence
  ))
} else {
  cat(
    "- This granularity's claims carry no separate confidence-qualifier column (whole-field claims, `complete_bm` only).\n"
  )
}
```

## Original BM vs. resulting claims

```{r}
#| label: split-view
#| results: asis
cat(x$html)
```
