---
title: NLI/LLM Funnel Report
subtitle: Narrowing down citing works that support or contradict a Background Message
date: today
params:
  assessment_id: "GA1"
  label: "REFUTES"
  granularity: "naive_bm"
  nli_active: "deberta_zeroshot"
format:
    html:
        toc: true
        toc-depth: 4
        embed-resources: true
        code-fold: true
        code-summary: 'Show the code'
        keep-md: false
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)

# Self-contained-report helper, same idiom as IPBES_Fact_Checker.qmd's setup
# chunk -- but report_files_dir is built from params$label/params$assessment_id/
# params$granularity, NOT knitr::current_input(). This qmd is rendered once
# per (assessment, label, granularity) combination via
# quarto::quarto_render(output_file=, execute_params=); current_input()
# always returns the SOURCE filename ("IPBES_Label_Funnel_Report"), which
# would collide every render onto the same _files/ directory.
#
# Inlined rather than sourced from R/branch_helpers.R's granularity_suffix()/
# nli_model_suffix(): quarto::quarto_render() runs this file in its own
# fresh R session, which doesn't inherit _targets.R's lapply(source) of
# R/*.R -- this qmd has never sourced project R/ helpers (only library()
# calls), so a one-line inline copy is more consistent than introducing
# that pattern here.
gran_suffix <- paste0("_", params$granularity)
model_suffix <- if (identical(params$nli_active, "deberta_zeroshot")) "" else paste0("_", params$nli_active)
# file.path("output/reports", ...) rather than a bare name: the qmd source
# lives in input/reports/, but the rendered .html gets moved to
# output/reports/ by the _targets.R target after this chunk runs (Quarto
# renders next to the input regardless of execute_dir) -- writing the
# sidecar directly into its final home means it's already correctly a
# sibling of the .html once that move happens.
report_files_dir <- file.path("output/reports", paste0("IPBES_", params$label, "_Report_", params$assessment_id, model_suffix, gran_suffix, "_files"))
# Wipe before repopulating: copy_into_report_files() only copies in what it's
# given, so a file dropped from one run's inputs (e.g. a removed funnel
# level) would otherwise linger here forever across re-renders.
unlink(report_files_dir, recursive = TRUE, force = TRUE)
dir.create(report_files_dir, recursive = TRUE, showWarnings = FALSE)
copy_into_report_files <- function(paths) {
  file.copy(paths, report_files_dir, overwrite = TRUE)
  file.path(basename(report_files_dir), basename(paths))
}

# Target names are label-specific (refutes_funnel_data / supports_funnel_data,
# etc.) rather than one generic target, so tar_read_raw() (the string-based
# variant) is used to pick the right one at render time from params$label.
# Matched by exact basename (not a loose grepl) against the deterministic
# filename the R builders produce -- necessary now that "naive_bm" (no
# suffix) and "complete_bm" ("_complete_bm" suffix) coexist on disk and a
# suffix-less grepl for e.g. "_GA1.rds" would also match "_GA1_complete_bm.rds"
# style names as a false substring hit in the wrong direction, or vice versa.
target_stem <- tolower(params$label)
expected_name <- function(kind, ext) sprintf("%s_funnel_%s_%s%s%s.%s", target_stem, kind, params$assessment_id, model_suffix, gran_suffix, ext)

data_path <- targets::tar_read_raw(paste0(target_stem, "_funnel_data")) |>
  (\(x) x[basename(x) == expected_name("data", "rds")])()
x <- readRDS(data_path)

fig_paths <- copy_into_report_files(
  targets::tar_read_raw(paste0(target_stem, "_funnel_figures")) |>
    (\(x) x[grepl(paste0("_", params$assessment_id, model_suffix, gran_suffix, "\\.png$"), x)])()
)
fig_overall <- fig_paths[grepl("_overall_", fig_paths)]
fig_by_bm <- fig_paths[grepl("_by_bm_", fig_paths) & !grepl("_normalized_", fig_paths)]
fig_by_bm_norm <- fig_paths[grepl("_by_bm_normalized_", fig_paths)]

table_paths <- copy_into_report_files(
  targets::tar_read_raw(paste0(target_stem, "_funnel_tables")) |>
    (\(x) x[grepl(paste0("_", params$assessment_id, model_suffix, gran_suffix, "\\.html$"), x)])()
)
table_l3 <- table_paths[grepl("_l3_", table_paths)]
```

# `r params$label` Funnel — `r params$assessment_id`

## Methodology

Every work discovered by the snowball search under a Background Message
(BM) is put through a 3-level sieve, each level a **subset** of the one
before it, counted as **distinct citing works per (assessment, km, bm)** —
a work cited under two different BMs is two separate findings, not one, so
per-BM counts sum to the assessment-wide total shown below.

1. **Snowball corpus** — every citing work discovered for that BM (the
   forward/citing direction of the snowball search).
2. **NLI `r params$label` (certain)** — of those, works with at least one
   claim scored `` label == "`r params$label`" `` with `uncertain == FALSE`
   by the zero-shot NLI model (Phase 1 — see
   [TD_BM_NLI_approach](TD_BM_NLI_approach.html)).
3. **LLM-confirmed `r params$label`** — of those, works where OpenRouter's
   LLM review (Phase 2 — see
   [TD_NLI_LLM_two_phase](TD_NLI_LLM_two_phase.html)) independently agreed
   with the `r params$label` verdict (`llm_agrees == TRUE`).

There is deliberately no further "+ sufficient evidence" level: Phase 2's
own verification logic forces its `llm_label` to `NOT_ENOUGH_INFO` whenever
it judges the evidence insufficient, so an LLM verdict of `r params$label`
already *implies* sufficient evidence — a separate level here would always
contain exactly the same works as level 3.

**NLI model: `r params$nli_active`.** Level 2/3 counts below reflect
whichever model was active in `input/config.yaml`'s `nli.active` when this
report was rendered — different models are not directly comparable
work-for-work either, since each has its own score distribution and
`uncertain_threshold` calibration (see `TD_BM_NLI_approach.qmd`).

**Claim granularity: `r params$granularity`.** `naive_bm` (the default)
segments each BM's `bm_description`/`bm_label` into evidence-delimited
sub-claims (2-17 per BM) at the sentence level. `complete_bm` instead
treats each field as a single, whole claim (2 per BM, or 1 if
`bm_label`/`bm_description` are identical) — cheaper to score (~71% fewer
NLI pairs measured across both assessments), at the cost of losing which
specific part of a longer BM a citing work actually supports or
contradicts, and of the resulting hypothesis text sometimes exceeding the
NLI model's own token limit (see `TD_BM_NLI_approach.qmd` for the measured
numbers and caveats). Level 2/3 counts below reflect whichever granularity
this report was rendered for; the two are **not** directly comparable
work-for-work, since they're scored against different claim text.

**Phase 2 (LLM verification) is a partial, ongoing process** — OpenRouter
credit has run out mid-run at least once already, so level 3 below
reflects only the pairs reviewed so far, not the full `r params$label`-certain
set from level 2. This number will grow as `llm_verification_parquet` gets
more OpenRouter runs; re-render this report after a rerun rather than
treating today's snapshot as final.

```{r}
#| label: empty-check
#| results: asis
if (isTRUE(x$empty)) {
  cat("No data available yet for this assessment — the snowball, NLI, or LLM verification stage hasn't produced output here.\n\n")
  knitr::knit_exit()
}
```

## Overall numbers

```{r}
#| label: overall-table
knitr::kable(
  x$funnel_overall,
  col.names = c("Level", "Description", "Distinct works")
)
```

```{r}
#| label: overall-figure
#| results: asis
if (length(fig_overall)) cat(sprintf("![](%s)\n\n", fig_overall))
```

## Per-BM numbers

```{r}
#| label: by-bm-table
knitr::kable(
  x$funnel_by_bm,
  col.names = c(
    "KM", "BM", "Corpus", sprintf("NLI %s", params$label), "LLM-confirmed",
    "% 2 of 1", "% 3 of 2"
  )
)
```

```{r}
#| label: by-bm-figure
#| results: asis
if (length(fig_by_bm)) cat(sprintf("![](%s)\n\n", fig_by_bm))
```

Normalized so every BM's own snowball corpus reads 1 (dashed line) — lets
BMs be compared on conversion rate regardless of raw corpus size.

```{r}
#| label: by-bm-figure-normalized
#| results: asis
if (length(fig_by_bm_norm)) cat(sprintf("![](%s)\n\n", fig_by_bm_norm))
```

## Level 3: LLM-confirmed `r params$label`

Every citing work where the LLM's own independent review agreed with NLI's
`r params$label` verdict (and, by construction, judged the evidence for it
sufficient — see Methodology). Use the BM dropdown in the filter row below
the header to narrow to one Background Message.

```{r}
#| label: level3-table
#| results: asis
if (length(table_l3)) {
  cat(sprintf(
    '<iframe src="%s" width="100%%" height="800px" style="border:none;"></iframe>\n\n',
    table_l3
  ))
}
```
