---
title: QA BM Split Report
subtitle: QA check for how each Background Message was split into claims
date: today
params:
  assessment_id: "GA1"
  granularity: "naive_bm"
format:
    html:
        toc: true
        toc-depth: 4
        embed-resources: true
        code-fold: true
        code-summary: 'Show the code'
        keep-md: false
---

```{r}
#| label: setup
#| include: false

library(dplyr)

# No report_files_dir/copy_into_report_files() here -- unlike the earlier
# DT::datatable() version of this report, the highlighted-text view is
# plain HTML embedded directly in this document (no separate widget file
# to ship as a sidecar), so there's nothing to copy in and no iframe/
# resource-path concerns at all.
gran_suffix <- paste0("_", params$granularity)

expected_name <- function(ext) {
  sprintf(
    "bm_split_highlighted_%s%s.%s",
    params$assessment_id,
    gran_suffix,
    ext
  )
}

data_path <- targets::tar_read_raw("bm_split_report_highlighted") |>
  (\(x) x[basename(x) == expected_name("rds")])()
x <- readRDS(data_path)

n_bm <- x$data |> distinct(km, bm) |> nrow()
n_claims <- nrow(x$data)
has_confidence <- "confidence" %in% names(x$data) && any(!is.na(x$data$confidence))
pct_with_confidence <- if (has_confidence) {
  round(100 * mean(!is.na(x$data$confidence)), 1)
} else {
  NA
}
```

# `r params$assessment_id` -- granularity: `r params$granularity`

## Methodology

Every Background Message (BM) in this assessment is split into claims by
`nli_ready_evidence_parquet` under the currently active `granularity`
setting (`input/config.yaml`'s `nli.active` config). This report is a
**QA check on that split, not a scoring result** — for every BM field, it
shows the original `bm_description`/`bm_label` text with each resulting
claim's source span colour-highlighted, followed by the claims themselves
as a colour-matched list, so the segmentation can be sanity-checked by eye
before any NLI/LLM scoring is run against it. Highlighting is based on the
raw span each claim was cut from, before any confidence-qualifier removal
or (for `atomic_bm`) LLM completion -- so it covers at least the major
part of a claim's origin, even where the listed claim text itself was
reworded. See `TD_BM_NLI_approach.qmd` for the full design rationale
behind each granularity option (`naive_bm` / `complete_bm` / `atomic_bm`).

```{r}
#| label: summary
#| results: asis
cat(sprintf("- **%d** distinct BMs, **%d** resulting claims\n", n_bm, n_claims))
if (has_confidence) {
  cat(sprintf(
    "- **%.1f%%** of claims had an IPBES confidence qualifier (\"well established\" / \"established but incomplete\" / \"unresolved\" / \"inconclusive\") extracted out of the claim text into its own column\n",
    pct_with_confidence
  ))
} else {
  cat(
    "- This granularity's claims carry no separate confidence-qualifier column (whole-field claims, `complete_bm` only).\n"
  )
}
```

## Original BM vs. resulting claims

```{r}
#| label: split-view
#| results: asis
cat(x$html)
```
