---
title: QA BM Split Report
subtitle: QA check for how each Background Message was split into claims
date: today
params:
  assessment_id: "GA1"
  granularity: "naive_bm"
format:
    html:
        toc: true
        toc-depth: 3
        embed-resources: true
        code-fold: true
        code-summary: 'Show the code'
        keep-md: false
---

```{r}
#| label: setup
#| include: false

library(dplyr)

# Self-contained-report helper, same idiom as IPBES_Label_Funnel_Report.qmd's
# setup chunk -- report_files_dir built from params$assessment_id/
# params$granularity, not knitr::current_input(), since this qmd is rendered
# once per assessment via quarto::quarto_render(output_file=, execute_params=).
#
# Inlined rather than sourced from R/branch_helpers.R's granularity_suffix():
# quarto::quarto_render() runs this file in its own fresh R session, which
# doesn't inherit _targets.R's lapply(source) of R/*.R -- this qmd has never
# sourced project R/ helpers (only library() calls), so a one-line inline
# copy is more consistent than introducing that pattern here.
gran_suffix <- paste0("_", params$granularity)
# file.path("output/reports", ...) rather than a bare name: the qmd source
# lives in input/reports/, but the rendered .html gets moved to
# output/reports/ by the _targets.R target after this chunk runs (Quarto
# renders next to the input regardless of execute_dir) -- writing the
# sidecar directly into its final home means it's already correctly a
# sibling of the .html once that move happens.
report_files_dir <- file.path("output/reports", paste0("QA_BM_Split_Report_", params$assessment_id, gran_suffix, "_files"))
unlink(report_files_dir, recursive = TRUE, force = TRUE)
dir.create(report_files_dir, recursive = TRUE, showWarnings = FALSE)
copy_into_report_files <- function(paths) {
  file.copy(paths, report_files_dir, overwrite = TRUE)
  file.path(basename(report_files_dir), basename(paths))
}

expected_name <- function(ext) sprintf("bm_split_report_table_%s%s.%s", params$assessment_id, gran_suffix, ext)

data_path <- targets::tar_read_raw("bm_split_report_table") |>
  (\(x) x[basename(x) == expected_name("rds")])()
x <- readRDS(data_path)

table_path <- copy_into_report_files(
  targets::tar_read_raw("bm_split_report_table") |>
    (\(x) x[basename(x) == expected_name("html")])()
)

n_bm <- x$data |> distinct(km, bm) |> nrow()
n_claims <- nrow(x$data)
has_confidence <- "confidence" %in% names(x$data)
pct_with_confidence <- if (has_confidence) round(100 * mean(!is.na(x$data$confidence)), 1) else NA
```

# `r params$assessment_id` -- granularity: `r params$granularity`

## Methodology

Every Background Message (BM) in this assessment is split into claims by
`nli_ready_evidence_parquet` under the currently active `granularity`
setting (`input/config.yaml`'s `nli.active` config). This report is a
**QA check on that split, not a scoring result** — it shows, for every
claim actually produced, the original `bm_description`/`bm_label` text it
came from, so the segmentation can be sanity-checked by eye before any
NLI/LLM scoring is run against it. See `TD_BM_NLI_approach.qmd` for the
full design rationale behind each granularity option (`naive_bm` /
`complete_bm` / `atomic_bm`).

```{r}
#| label: summary
#| results: asis
cat(sprintf("- **%d** distinct BMs, **%d** resulting claims\n", n_bm, n_claims))
if (has_confidence) {
  cat(sprintf(
    "- **%.1f%%** of claims had an IPBES confidence qualifier (\"well established\" / \"established but incomplete\" / \"unresolved\" / \"inconclusive\") extracted out of the claim text into its own column\n",
    pct_with_confidence
  ))
} else {
  cat("- This granularity's claims carry no separate confidence-qualifier column (whole-field claims, `complete_bm` only).\n")
}
```

## Original BM vs. resulting claims

Use the filter row below the header to narrow to one KM/BM/source field.

```{r}
#| label: split-table
#| results: asis
cat(sprintf(
  '<iframe src="%s" width="100%%" height="800px" style="border:none;"></iframe>\n\n',
  table_path
))
```
