---
title: "QA: NLI Model Benchmark"
subtitle: Held-out evaluation across the zero-shot baseline and every fine-tuned checkpoint
date: today
format:
    html:
        toc-depth: 5
        toc_expand: true
---

```{r}
#| label: targets-dependencies
#| include: false

# Named literally so tar_quarto() turns this into a real dependency edge.
# Hand-written rather than generated from config.yaml's `reports:` block:
# there is no dimension to expand over -- one report covers every model.
qa <- readRDS(targets::tar_read(nli_benchmark_qa_data))

library(dplyr)
library(ggplot2)
library(knitr)

pct <- function(x) ifelse(is.na(x), "--", sprintf("%.1f%%", 100 * x))
```

## What this measures, and what it does not

**It measures agreement with `gpt-4o-mini`.** Every label in the holdout comes
from Phase 2's grounded, quote-checked LLM verdict. Fine-tuning distils that
judgment into the NLI model and this report scores how well the distillation
worked. It does not measure truth: the LLM's verdicts have never been checked
against human labels. That is a ceiling, not a flaw — distilling an expensive
accurate judge into a cheap one is the point — but every number below inherits
it.

Design and reasoning: [TD_NLI_training](TD_NLI_training.html).

```{r}
#| label: fold
#| echo: false
f <- qa$fold
kable(data.frame(
  Property = c("benchmark salt", "holdout fraction", "grouping key",
               "train rows", "test rows", "train groups", "test groups",
               "claims in BOTH folds", "works in both folds",
               "test rows involving a shared work"),
  Value = c(f$salt, f$holdout_fraction, f$group_key,
            f$n_train, f$n_test, f$n_groups_train, f$n_groups_test,
            f$claims_in_both_folds,
            sprintf("%d of %d (%s)", f$works_in_both_folds, f$works_total,
                    pct(f$works_in_both_folds / f$works_total)),
            sprintf("%d (%s)", f$test_rows_with_shared_work,
                    pct(f$test_rows_with_shared_work / f$n_test)))
), caption = "Holdout composition and leakage diagnostics")
```

**`claims in BOTH folds` must be 0.** The holdout is grouped by
`(assessment, bm)` — deliberately *without* `km`, because one Background
Message belongs to several Key Messages and carries byte-identical claim text
under each. Grouping on `(assessment, km, bm)` looked tighter and was measurably
leakier: 141 groups, 126 claim strings present in both folds. If this number is
ever non-zero, the grouping key has drifted.

**Work overlap is reported, not eliminated.** A work cited under several BMs can
appear on both sides paired with *different* claims. That is much weaker than
claim-text leakage — the model sees the premise, never the pair or its label —
and closing it needs a connected-component split that at this size yields a few
huge unbalanced components, trading a small measurable bias for a large
unmeasurable one. Judge the headline numbers with the percentage above in mind.

## Models under test

```{r}
#| label: models
#| echo: false
kable(qa$models |> select(model_id, mode, split_aware, max_length, n_rows, n_truncated),
      caption = "Scored models. `zeroshot` = 3 forward passes per row with the deployed hypothesis template, entailment logits cross-normalised. `direct` = the fine-tuned head's own 3-way softmax. `split_aware` is NA for the untrained baseline.")
```

::: {.callout-important}
## `split_aware = FALSE` means the scores below are contaminated

A checkpoint trained **before** the holdout existed was trained on rows drawn
from the whole pool — including what is now the test fold. Its score here is
recall on memorised rows, not generalisation.

This is not hypothetical. The larger pre-split run trained on 558 of the
corpus's 562 REFUTES rows, so roughly **169 of the test fold's 170 REFUTES were
in its training data**, and it duly posts the best numbers in every table below.

Those runs are shown rather than hidden, because they are the clearest available
evidence for *how badly a leaking split flatters a model* — which is the finding
that motivated this benchmark. But **no table below is a ranking**. Compare only
rows where `split_aware` is `TRUE` against the baseline.
:::

One asymmetry that runs the other way, and makes the honest result
conservative: the fine-tuned models are scored at the 512 tokens they were
trained on, and 141 of 467 premises (30%) hit that limit, against 3 for the
baseline at its deployed 2048. The fine-tuned models are working from truncated
abstracts where the baseline is not.

The baseline is scored the way it is actually **deployed** — a binary
entailment head cannot emit SUPPORTS/REFUTES/NEI directly, so each candidate
label is folded into its own reformulated hypothesis and the three entailment
logits are cross-normalised. A baseline scored differently from how it runs in
production is not a baseline.

One caveat on exactness: the deployed pipeline cleans the premise
(`clean_title()`/`clean_abstract()` strip HTML/JATS/LaTeX) while this benchmark
uses raw `title`/`abstract`, matching what `train_nli.py` trains on. Spot-checked
against real deployed scores, local and production probabilities agree to ~4
decimal places on clean abstracts and diverge on marked-up ones.

## The headline: SUPPORTS → REFUTES recovery

60% of every confirmed refutation in this corpus is a pair the zero-shot model
called **SUPPORTS** and the LLM overturned. It did not merely miss them — it got
them backwards. So the question that decides whether fine-tuning did this
project's job is: of the holdout rows where NLI said SUPPORTS and the truth is
REFUTES, how many does each model now call REFUTES?

```{r}
#| label: flip
#| echo: false
qa$flip |>
  left_join(qa$models |> select(model, model_id, split_aware), by = "model") |>
  transmute(model_id, split_aware, n_subset, n_recovered, recovery_rate = pct(recovery_rate)) |>
  arrange(desc(n_recovered)) |>
  kable(caption = "Rows where NLI said SUPPORTS and the truth is REFUTES. Read only the split_aware = TRUE rows against the baseline (NA).")
```

A checkpoint that does not move this number has not earned deployment, whatever
its macro-F1 says.

## Per-class performance

Accuracy is deliberately absent. The holdout is near-balanced while deployment
is 0.2–3% REFUTES, so a balanced-set accuracy flatters every model equally and
predicts nothing about the real stream. REFUTES precision and recall are given
separately and never folded into F1 for the headline: a missed refutation is a
finding that never surfaces, a false one is a claim a human discards in seconds,
and F1 assumes those cost the same.

```{r}
#| label: per-class
#| echo: false
qa$per_class |>
  left_join(qa$models |> select(model, model_id, split_aware), by = "model") |>
  transmute(model_id, split_aware, label, precision = pct(precision), recall = pct(recall),
            f1 = pct(f1), support) |>
  arrange(label, model_id) |>
  kable()
```

```{r}
#| label: confusion
#| echo: false
qa$confusion |>
  left_join(qa$models |> select(model, model_id), by = "model") |>
  tidyr::pivot_wider(id_cols = c(model_id, label), names_from = pred_label,
                     values_from = n, values_fill = 0) |>
  arrange(model_id, label) |>
  kable(caption = "Confusion matrix: rows = LLM verdict (truth), columns = model prediction")
```

## Threshold sweep

Phase 1 stores the full probability distribution and every threshold decision is
made downstream in `dplyr`, so a single argmax operating point would describe
neither the model nor the deployment. Framed as a REFUTES detector — predict
REFUTES iff `p_refutes >= t` — which is what these scores are actually used for.
This is also where the re-calibrated deployment threshold comes from: the
base-rate shift between a balanced training fold and a 0.2% deployment stream
guarantees an argmax classifier over-predicts REFUTES in production.

```{r}
#| label: sweep
#| echo: false
#| fig-width: 9
#| fig-height: 5
qa$threshold_sweep |>
  left_join(qa$models |> select(model, model_id), by = "model") |>
  tidyr::pivot_longer(c(precision, recall), names_to = "metric", values_to = "value") |>
  ggplot(aes(threshold, value, colour = model_id, linetype = metric)) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "p_refutes threshold", y = NULL, colour = "model", linetype = NULL,
       title = "REFUTES detection: precision and recall across thresholds") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box = "vertical")
```

## Per assessment

558 of 562 REFUTES in the pool come from GA1 and IAS, so a pooled number mostly
measures those two. TCA and VA are shown separately even though their counts are
tiny — "we cannot tell" is a legitimate and useful result, and whether anything
learned here generalises beyond GA1/IAS is an open question.

```{r}
#| label: by-assessment
#| echo: false
qa$by_assessment |>
  left_join(qa$models |> select(model, model_id), by = "model") |>
  transmute(model_id, assessment, n, n_refutes,
            refutes_recall = pct(refutes_recall),
            refutes_precision = pct(refutes_precision)) |>
  arrange(assessment, model_id) |>
  kable()
```

## What this cannot tell you

**Recall against the real corpus.** The holdout inherits Phase 2's routing bias:
for citing works only `REFUTES-certain` and `SUPPORTS-certain` pairs were ever
sent for LLM review, so refutations hiding among the 1.23M `NOT_ENOUGH_INFO`
pairs in GA1 alone are in no fold of any split. The full-coverage key-paper chain
estimates 0.37% of NEI pairs are genuine refutations against 3.09% of REFUTES
pairs — an 8x enrichment across a 4.8x larger pile, meaning **37% of the
refutations found there were sitting in the NEI bucket**. Every number above is
conditional on a pair having been routed for review in the first place.

Closing that needs a recall probe: a random sample across the full NLI output
including NEI, labelled by Phase 2. See [TD_NLI_training](TD_NLI_training.html).
